---
title: "CorNetto Workflow"
output:
  BiocStyle::html_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{CorNetto Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(CorNetto)
```

# Introduction

CorNetto builds knowledge-guided multi-omic correlation networks from
normalized assay data stored in a `MultiAssayExperiment`. It is intended
for workflows where transcriptomic, proteomic, metabolomic, or related
assays are already normalized and the analysis question is whether
groups differ in correlation structure, prior-supported connectivity, or
node-level rewiring.

The package uses standard Bioconductor containers for assay and sample
metadata, and returns standardized edge tables that can be combined,
converted to `igraph` objects, or exported as Cytoscape-ready node and
edge tables.

# Installation

After acceptance to Bioconductor, install CorNetto with:

```{r install, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}
BiocManager::install("CorNetto")
```

# Example data

```{r load-data}
analysisData <- exampleAnalysisData()
knowledgeNetwork <- exampleKnowledgeNetwork()
seedNodes <- exampleSeedNodes()

analysisData
head(knowledgeNetwork)
seedNodes
```

The synthetic example contains three assays named `protein`,
`transcript`, and `metabolite`. The example prior network also contains a
drug-target edge to demonstrate that prior networks can include static
knowledge beyond the measured assays. Sparse correlation functions must
therefore be given only prior edges whose assay names are present in the
analysis object.

```{r measured-priors}
measuredAssays <- c("protein", "transcript", "metabolite")
measuredKnowledgeNetwork <- knowledgeNetwork[
    knowledgeNetwork$fromAssayName %in% measuredAssays &
        knowledgeNetwork$toAssayName %in% measuredAssays,
    ,
    drop = FALSE
]
```

# Dense group-specific correlations

`createCorrelationNetwork()` computes all pairwise within-omic
correlations for one assay and one group. `createCorrelationNetworks()`
is the multi-assay or multi-group wrapper.

```{r dense-correlations}
recoveredProteinNetwork <- createCorrelationNetwork(
    analysisData = analysisData,
    assayName = "protein",
    groupColumn = "clinicalGroup",
    groupLevel = "Recovered",
    correlationMethod = "pearson",
    minimumAbsoluteCorrelation = 0,
    adjustedPValueThreshold = 1,
    pAdjustMethod = "fdr",
    storeResult = FALSE
)

head(recoveredProteinNetwork)
```

# Prior-guided sparse correlations

Sparse correlation functions calculate only feature pairs supported by a
prior network. This avoids exhaustive cross-omic testing and keeps the
dynamic network aligned to interpretable prior knowledge.

```{r sparse-correlations}
sparseMulti <- createSparseMultiOmicCorrelations(
    analysisData = analysisData,
    knowledgeNetwork = measuredKnowledgeNetwork,
    correlationScope = "all",
    groupColumn = "clinicalGroup",
    groupLevel = "Recovered",
    minimumAbsoluteCorrelation = 0,
    adjustedPValueThreshold = 1,
    storeResult = FALSE
)

combinedDynamicNetwork <- combineCorrelationNetworks(
    recoveredProteinNetwork,
    sparseMulti
)

c(nrow(sparseMulti), nrow(combinedDynamicNetwork))
```

# Permutation validation

`validateSparseCorrelationEdges()` estimates empirical p-values for
prior-supported sparse correlations. `permuteDifferentialRewiring()`
permutes group labels and reruns the differential-correlation workflow
to validate node-level rewiring scores.

```{r permutation-validation}
sparseValidation <- validateSparseCorrelationEdges(
    analysisData = analysisData,
    knowledgeNetwork = measuredKnowledgeNetwork,
    correlationScope = "all",
    groupColumn = "clinicalGroup",
    groupLevel = "Recovered",
    nPermutations = 10,
    seed = 1
)

head(sparseValidation)
```

# Differential correlation and rewiring

`testDifferentialCorrelation()` compares correlation coefficients between
two groups. `createDifferentialCorrelationNetwork()` converts the test
results to a weighted network, and `calculateRewiringScores()` summarizes
node-level rewiring.

```{r differential-correlation}
differentialResults <- testDifferentialCorrelation(
    analysisData = analysisData,
    groupColumn = "clinicalGroup",
    groupLevels = c("PASC", "Recovered"),
    assayName = "protein",
    correlationMethod = "pearson",
    minimumAbsoluteCorrelation = 0,
    adjustedPValueThreshold = 1,
    pAdjustMethod = "fdr",
    storeResult = FALSE
)

differentialNetwork <- createDifferentialCorrelationNetwork(
    differentialCorrelationTable = differentialResults,
    differenceAdjustedPValueThreshold = 1,
    minimumAbsoluteCorrelation = 0
)

rewiringScores <- calculateRewiringScores(
    differentialCorrelationNetwork = differentialNetwork,
    storeResult = FALSE
)

head(rewiringScores)
```

# Integrated and focused network

`createIntegratedNetwork()` merges static prior knowledge and dynamic
correlation layers. `createFocusedNetwork()` extracts a seed-centered
neighborhood, which is useful for pathway-driven analyses.

```{r integrated-network}
integratedNetwork <- createIntegratedNetwork(
    knowledgeNetwork = knowledgeNetwork,
    correlationNetwork = combinedDynamicNetwork,
    differentialCorrelationNetwork = differentialNetwork,
    includeReverseEdges = TRUE
)

focusedNetwork <- createFocusedNetwork(
    networkEdgeTable = integratedNetwork,
    seedNodes = seedNodes,
    neighborhoodOrder = 1
)

cytoscapeTables <- prepareCytoscapeTables(
    networkEdgeTable = focusedNetwork,
    rewiringTable = rewiringScores
)

names(cytoscapeTables)
nrow(cytoscapeTables$nodes)
nrow(cytoscapeTables$edges)
```

# Graph creation

`createNetworkGraph()` converts a standardized CorNetto edge table to an
`igraph` object. Plotting can then use standard igraph methods.

```{r graph-plot, fig.width = 8, fig.height = 6}
graphObject <- createNetworkGraph(focusedNetwork)
plot(
    graphObject,
    layout = igraph::layout_with_fr(graphObject),
    vertex.label = NA,
    vertex.size = 8,
    edge.width = 1,
    main = "Example focused CorNetto network"
)
```

# Export

Use `writeNetworkTables()` when Cytoscape-ready files are needed. The
example writes to a temporary directory.

```{r export}
writtenFiles <- writeNetworkTables(
    networkTables = cytoscapeTables,
    directoryPath = tempdir(),
    prefix = "cornettoExample",
    fileFormat = "tsv"
)

basename(writtenFiles)
```

# Session information

```{r session-info}
sessionInfo()
```
