---
title: "CorNetto Workflow"
author: "Bradley Ward"
date: "2026-05-19"
output:
  BiocStyle::html_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{CorNetto Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(CorNetto)
```

# Introduction

CorNetto builds knowledge-guided multi-omic correlation networks from
normalized assay data stored in a `MultiAssayExperiment`. It is intended
for workflows where transcriptomic, proteomic, metabolomic, or related
assays are already normalized and the analysis question is whether
groups differ in correlation structure, prior-supported connectivity, or
node-level rewiring.

The package uses standard Bioconductor containers for assay and sample
metadata, and returns standardized edge tables that can be combined,
converted to `igraph` objects, or exported as Cytoscape-ready node and
edge tables.

CorNetto is for downstream analysis after assay-level preprocessing 
by reusing Bioconductor containers and established infrastructure such 
as `DGCA` for dense differentialcorrelation testing and `igraph` for 
graph representation. The packageadds the glue needed for prior-guided 
candidate-edge differential correlation, knowledge-network-aware integration, 
focused subnetworks, and rewiring summaries in a single reproducible workflow.

# Version information

```{r version-information}
c(
    R = R.version.string,
    Bioconductor = as.character(BiocManager::version()),
    CorNetto = as.character(utils::packageVersion("CorNetto"))
)
```

# Installation

After acceptance to Bioconductor, install CorNetto with:

```{r install, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}
BiocManager::install("CorNetto")
```

# Example data

```{r load-data}
analysisData <- exampleAnalysisData()
knowledgeNetwork <- exampleKnowledgeNetwork()
seedNodes <- exampleSeedNodes()

analysisData
summarizeAnalysisData(analysisData, groupColumn = "clinicalGroup", quiet = TRUE)
head(knowledgeNetwork)
seedNodes
```

The synthetic example contains three assays named `protein`,
`transcript`, and `metabolite`. The example prior network also contains a
drug-target edge to demonstrate that prior networks can include static
knowledge beyond the measured assays. Prior-guided candidate-edge
analyses must therefore be given only prior edges whose assay names are
present in the analysis object.

```{r measured-priors}
measuredAssays <- c("protein", "transcript", "metabolite")
measuredKnowledgeNetwork <- knowledgeNetwork[
    knowledgeNetwork$fromAssayName %in% measuredAssays &
        knowledgeNetwork$toAssayName %in% measuredAssays,
    ,
    drop = FALSE
]
```

# Dense group-specific correlations

`createCorrelationNetwork()` computes all pairwise within-omic
correlations for one assay and one group. Dense all-pairs mode is
intended for small-to-medium assays; for large assays, prefer
prior-guided candidate-edge testing when suitable.

```{r dense-correlations}
recoveredProteinNetwork <- createCorrelationNetwork(
    analysisData = analysisData,
    assayName = "protein",
    groupColumn = "clinicalGroup",
    groupLevel = "Recovered",
    correlationMethod = "pearson",
    minimumAbsoluteCorrelation = 0,
    adjustedPValueThreshold = 1,
    pAdjustMethod = "fdr",
    storeResult = FALSE
)

head(recoveredProteinNetwork)
```

# Prior-guided sparse differential correlation

`testDifferentialCorrelation()` can test only feature pairs supported by
a prior network when `candidateEdgeTable` is supplied. This is the sparse
mode for differential correlation: it avoids exhaustive cross-omic
testing and keeps the dynamic network aligned to interpretable prior
knowledge.

```{r prior-guided-differential}
priorGuidedDifferentialResults <- testDifferentialCorrelation(
    analysisData = analysisData,
    candidateEdgeTable = measuredKnowledgeNetwork,
    groupColumn = "clinicalGroup",
    groupLevels = c("PASC", "Recovered"),
    minimumAbsoluteCorrelation = 0,
    adjustedPValueThreshold = 1,
    pAdjustMethod = "fdr",
    storeResult = FALSE
)

priorGuidedDifferentialNetwork <- createDifferentialCorrelationNetwork(
    differentialCorrelationTable = priorGuidedDifferentialResults,
    differenceAdjustedPValueThreshold = 1,
    minimumAbsoluteCorrelation = 0
)

c(nrow(priorGuidedDifferentialResults), nrow(priorGuidedDifferentialNetwork))
```

# Permutation validation

`permuteRewiringScores()` permutes group labels and reruns the
differential-correlation workflow to validate node-level rewiring
scores.

```{r permutation-validation}
rewiringValidation <- permuteRewiringScores(
    analysisData = analysisData,
    candidateEdgeTable = measuredKnowledgeNetwork,
    groupColumn = "clinicalGroup",
    groupLevels = c("PASC", "Recovered"),
    minimumAbsoluteCorrelation = 0,
    adjustedPValueThreshold = 1,
    differenceAdjustedPValueThreshold = 1,
    pAdjustMethod = "fdr",
    nPermutations = 2,
    seed = 1
)

head(rewiringValidation$rewiringTable)
```

# Differential correlation and rewiring

`testDifferentialCorrelation()` compares correlation coefficients between
two groups. `createDifferentialCorrelationNetwork()` converts the test
results to a weighted network, and `calculateRewiringScores()` summarizes
node-level rewiring.

```{r differential-correlation}
differentialResults <- testDifferentialCorrelation(
    analysisData = analysisData,
    groupColumn = "clinicalGroup",
    groupLevels = c("PASC", "Recovered"),
    assayName = "protein",
    correlationMethod = "pearson",
    minimumAbsoluteCorrelation = 0,
    adjustedPValueThreshold = 1,
    pAdjustMethod = "fdr",
    storeResult = FALSE
)

differentialNetwork <- createDifferentialCorrelationNetwork(
    differentialCorrelationTable = differentialResults,
    differenceAdjustedPValueThreshold = 1,
    minimumAbsoluteCorrelation = 0
)

rewiringScores <- calculateRewiringScores(
    differentialCorrelationNetwork = differentialNetwork,
    storeResult = FALSE
)

head(rewiringScores)
```

# Integrated and focused network

`combineNetworks()` merges static prior knowledge,
group-specific correlation layers, and differential-correlation layers.
`createFocusedNetwork()` extracts a seed-centered neighborhood, which is
useful for pathway-driven analyses.

```{r integrated-network}
integratedNetwork <- combineNetworks(
    knowledgeNetwork = knowledgeNetwork,
    correlationNetwork = recoveredProteinNetwork,
    differentialCorrelationNetwork = list(
        priorGuidedDifferentialNetwork,
        differentialNetwork
    ),
    includeReverseEdges = TRUE
)

focusedNetwork <- createFocusedNetwork(
    networkEdgeTable = integratedNetwork,
    seedNodes = seedNodes,
    neighborhoodOrder = 1
)

cytoscapeTables <- prepareCytoscapeTables(
    networkEdgeTable = focusedNetwork,
    rewiringTable = rewiringScores
)

names(cytoscapeTables)
nrow(cytoscapeTables$nodes)
nrow(cytoscapeTables$edges)
```

# Graph creation

`createNetworkGraph()` converts a standardized CorNetto edge table to an
`igraph` object. Plotting can then use standard igraph methods.

```{r graph-plot, fig.width = 8, fig.height = 6}
graphObject <- createNetworkGraph(focusedNetwork)
plot(
    graphObject,
    layout = igraph::layout_with_fr(graphObject),
    vertex.label = NA,
    vertex.size = 8,
    edge.width = 1,
    main = "Example focused CorNetto network"
)
```

# Export

Use `writeNetworkTables()` when Cytoscape-ready files are needed. The
example writes to a temporary directory.

```{r export}
writtenFiles <- writeNetworkTables(
    networkTables = cytoscapeTables,
    directoryPath = tempdir(),
    prefix = "cornettoExample",
    fileFormat = "tsv"
)

basename(writtenFiles)
```

# Session information

```{r session-info}
sessionInfo()
```
