---
title: "CorNetto COVID-19 Severity Workflow"
author: "Bradley Ward"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{CorNetto COVID Severity Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(CorNetto)
extdataDir <- system.file("extdata", package = "CorNetto")
```

# Introduction

CorNetto builds knowledge-guided multi-omic correlation networks from
normalized assay data stored in a `MultiAssayExperiment`. It is intended
for workflows where transcriptomic, proteomic, metabolomic, or related
assays are already normalized and the analysis question is whether groups
differ in correlation structure, prior-supported connectivity, or
node-level rewiring.

The package uses standard Bioconductor containers for assay and sample
metadata, and returns standardized edge tables that can be combined,
converted to `igraph` objects, or exported as Cytoscape-ready node and
edge tables.

This vignette uses a small packaged subset of proteomic, transcriptomic,
and metabolomic data from moderate and severe COVID-19 patients. The subset
is intended to demonstrate the workflow during package checks; it is not
sized for biological discovery. The candidate edges used below are entirely
synthetic. They demonstrate network import, filtering, differential
correlation, and node scoring, but they do not represent biological evidence.

Constraining the calculations to an explicit candidate network keeps this
example fast and avoids dense all-pairs correlations.

# Version information

```{r version-information}
c(
    R = R.version.string,
    Bioconductor = as.character(BiocManager::version()),
    CorNetto = as.character(utils::packageVersion("CorNetto"))
)
```

# Installation

After acceptance to Bioconductor, install CorNetto with:

```{r install, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}
BiocManager::install("CorNetto")
```

# Import synthetic candidate networks

CorNetto accepts prior-knowledge networks from external resources when their
columns can be mapped to the package schema. For this runnable example, the
package instead supplies three deterministic synthetic networks. This avoids
redistributing third-party interaction data and keeps the example independent
of database versions and licenses.

The within-assay and cross-assay files contain invented edges between measured
features. The decoy file contains invented edges with one or two unmeasured
endpoints, so the filtering step below has observable work to do. Edge weights
and directions are illustrative only.

`readKnowledgeNetwork()` function takes a delimited table, optionally remaps
the input column names to a common convention, and standardizes it for downstream
CorNetto functions. The minimum columns required are `fromFeatureIdentifier`,
`toFeatureIdentifier`, `fromAssayName`, and `toAssayName`. Other columns are
optional but can be useful. The output is a standardized `S4Vectors::DataFrame`.

`combineKnowledgeNetworks()` takes these standardized networks and combines their rows while removing duplicate edges (optional).

```{r load prior data}

# Set column mapping
priorColumnMap <- c(
    fromFeatureIdentifier = "from",
    toFeatureIdentifier = "to",
    fromFeatureName = "fromName",
    toFeatureName = "toName",
    fromAssayName = "fromOmic",
    toAssayName = "toOmic",
    edgeType = "edgeType",
    edgeDirection = "edgeDirection",
    evidenceScore = "edgeWeight"
)
# Import the three synthetic edge sets
withinAssayNetwork <- readKnowledgeNetwork(
    filePath = file.path(
        extdataDir, "priorNetworks", "synthetic_within_assay.csv"
    ),
    columnMapping = priorColumnMap,
    knowledgeSource = "CorNetto synthetic within-assay"
)

crossAssayNetwork <- readKnowledgeNetwork(
    filePath = file.path(
        extdataDir, "priorNetworks", "synthetic_cross_assay.csv"
    ),
    columnMapping = priorColumnMap,
    knowledgeSource = "CorNetto synthetic cross-assay"
)

decoyNetwork <- readKnowledgeNetwork(
    filePath = file.path(
        extdataDir, "priorNetworks", "synthetic_decoy_edges.csv"
    ),
    columnMapping = priorColumnMap,
    knowledgeSource = "CorNetto synthetic decoys"
)

# Combine networks
combinedPriorNetwork <- combineKnowledgeNetworks(
    withinAssayNetwork,
    crossAssayNetwork,
    decoyNetwork,
    removeDuplicates = TRUE
)
```

# Import experimental abundance data into the environment

Currently CorNetto is able to accept any normalised and fully preprocessed
abundance or count matrix relating to proteomics, transcriptomics, or metabolomics.

`createAnalysisData()` creates a MultiAssayExperiment from these normalized assay objects and sample metadata.

`filterSamples()` filters the MultiAssayExperiment based on a group column and a range of values.

```{r load experimental data}
# Import patient data
patientInformation <- read.csv(file.path(extdataDir,
                                         "covidData",
                                         "patientInformation.csv"))
rownames(patientInformation) <- patientInformation$Visit_ID

# Import omics data
dataList <- list(RNA = read.csv(file.path(extdataDir,
                                          "covidData",
                                          "rnaMatrix.csv"),
                                row.names = 1),
                 Protein  = read.csv(file.path(extdataDir,
                                               "covidData",
                                               "proteinMatrix.csv"),
                                     row.names = 1),
                 Metabolite  = read.csv(file.path(extdataDir,
                                                  "covidData",
                                                  "metaboliteMatrix.csv"),
                                        row.names = 1))

# Create analysis data
analysisData <- createAnalysisData(assayList = dataList,
                                   sampleData = patientInformation)

# Filter data for only COVID-19 patients at visit 1
analysisCovid <- filterSamples(analysisData,
                               groupColumn = "Group",
                               groupLevels = c("COVID Severe",
                                               "COVID Moderate"))
analysisCovid <- filterSamples(analysisCovid,
                               groupColumn = "Visit",
                               groupLevels = "Visit 1")
```

# QC checks of all imported objects

Following data importation, objects can be checked for structural conformity and
diagnostic summaries.

`validateAnalysisData()` checks the CorNetto analysis object for structural conformity, specifically:

* is this a MultiAssayExperiment?
* does it contain assays?
* are sample identifiers valid and unique?
* are assay columns matched to colData?
* are feature/sample names present and non-duplicated?
* are assays extractable as numeric matrices?
* are required feature metadata fields present?

`validateKnowledgeNetwork()` checks the CorNetto prior-knowledge edge table
for structural conformity, specifically:

* is this a data frame-like edge table?
* are the required edge columns present?
* are source and target feature identifiers present?
* are source and target assay names present?
* are missing optional columns filled with standard defaults?
* are edge type, edge direction, and knowledge source represented consistently?
* is the network converted to the standardized CorNetto edge-table format?

These functions also run quietly after most CorNetto functions to ensure
conformity of output.

```{r validateAnalysisData}
# Check analysis data
analysisCovid <- validateAnalysisData(analysisCovid)
# Check knowledge networks
combinedPriorNetwork <- validateKnowledgeNetwork(combinedPriorNetwork)
```

`summarizeAnalysisData()` asks "what does this analysis object look like,
and are there analysis risks?" rather than only asking whether the object
is structurally valid.
```{r summarizeAnalysisData}
summarizeAnalysisData(analysisCovid,
                      groupColumn = "Group")
```

The output reports the number of features per assay, the number of samples
per assay, and the number of samples in each selected group. In this small
packaged example there are 12 moderate COVID-19 samples and 12 severe
COVID-19 samples.

Zero-variance and near-zero-variance features should be removed before
correlation analysis. These can be removed with `removeZeroVariance = TRUE`
and `minimumVariance` in `filterFeatures()`.

```{r filterFeatures}
analysisCovid <- filterFeatures(
    analysisCovid,
    removeZeroVariance = TRUE,
    minimumVariance = 0.01
)

summarizeAnalysisData(analysisCovid)
```

For the sparse workflow, candidate edges are restricted to edges whose
endpoint features remain in the filtered analysis object. The first summary
shows that the combined synthetic network includes deliberately unmeasured
endpoints. The second summary shows the network after requiring both endpoints
to be measured.

```{r restrict-prior-network}
measuredNodeKeys <- unlist(
    lapply(
        names(MultiAssayExperiment::experiments(analysisCovid)),
        function(assayName) {
            assayObject <- MultiAssayExperiment::experiments(analysisCovid)[[assayName]]
            paste(
                assayName,
                rownames(SummarizedExperiment::assay(assayObject)),
                sep = "::"
            )
        }
    ),
    use.names = FALSE
)

unfilteredPriorSummary <- summarizeKnowledgeNetwork(
    combinedPriorNetwork,
    analysisCovid,
    quiet = TRUE
)
unfilteredPriorSummary$overall

measuredPriorNetwork <- filterNetworkByNodes(
    combinedPriorNetwork,
    nodes = measuredNodeKeys,
    mode = "both",
    nodeIdType = "key"
)

measuredPriorSummary <- summarizeKnowledgeNetwork(
    measuredPriorNetwork,
    analysisCovid,
    quiet = TRUE
)
measuredPriorSummary$overall
```

# Run prior-guided differential correlations

The main function for testing differential correlations between groups is
`testDifferentialCorrelation`. Here it tests every measured edge in the
synthetic candidate network supplied through `candidateEdgeTable`. The result
illustrates the software workflow and is not a prior-supported biological
analysis.

The differential correlation result is added to `analysisCovid`, then
extracted with `differentialCorrelationResults()`. Alternatively,
`testDifferentialCorrelation()` can directly return the table with
`storeResult = FALSE`.
```{r testDifferentialCorrelation}
analysisCovid <- testDifferentialCorrelation(
    analysisData = analysisCovid,
    groupColumn = "Group",
    groupLevels = c("COVID Moderate", "COVID Severe"),
    candidateEdgeTable = measuredPriorNetwork,
    correlationMethod = "pearson",
    minimumAbsoluteCorrelation = 0.3,
    adjustedPValueThreshold = 0.05,
    pAdjustMethod = "fdr",
    resultName = "differentialSparseCorrelations"
)

differentialSparseCorrelations <-
    differentialCorrelationResults(analysisCovid)[["differentialSparseCorrelations"]]

as.data.frame(differentialSparseCorrelations)

```

# Build differential network.

The next step builds a differential correlation network from the
`differentialSparseCorrelations` table. The function returns the network
directly unless `storeResult = TRUE`, in which case `analysisData` and
`resultName` are supplied.

Here the resulting network is not filtered further. In a real analysis,
`minimumAbsoluteCorrelation` and `differenceAdjustedPValueThreshold` can
be tightened to remove weakly correlated or non-significant edges.
```{r}
analysisCovid <- createDifferentialCorrelationNetwork(
    differentialCorrelationTable = differentialSparseCorrelations,
    minimumAbsoluteCorrelation = 0,
    edgeWeightMethod = "signedZScore",
    analysisData = analysisCovid,
    resultName = "differentialSparseNetwork",
    storeResult = TRUE)

differentialSparseNetwork <-
    differentialCorrelationNetworks(analysisCovid)[["differentialSparseNetwork"]]

as.data.frame(differentialSparseNetwork)

```

# Calculate rewiring scores.

Rewiring scores summarize the amount of differential-correlation signal
incident on each node in the differential network. Here the rewiring table
is stored in `analysisCovid` and extracted with `rewiringResults()`.

```{r calculateRewiringScores}
analysisCovid <- calculateRewiringScores(
    differentialCorrelationNetwork = differentialSparseNetwork,
    analysisData = analysisCovid,
    resultName = "differentialSparseRewiring",
    storeResult = TRUE
)

rewiringScores <- rewiringResults(analysisCovid)[["differentialSparseRewiring"]]

rewiringScores <- rewiringScores[
    order(rewiringScores$rootMeanSquareRewiringScore, decreasing = TRUE),
]

head(rewiringScores, 20)
```

# Run permutation ranking.

`permuteRewiringScores()` permutes the supplied group labels, reruns the
differential correlation testing, rebuilds the networks, and calculates
node-level permutation tail probabilities. This example retains the observed
correlation and adjusted-p filters used above, so its output is a conditional
ranking, not a randomization p-value. This vignette uses 99 Monte Carlo
permutations so that package checks run quickly. Assignments are sampled
independently and can repeat; exact enumeration is not implemented.

```{r}
analysisCovid <- permuteRewiringScores(
    analysisData = analysisCovid,
    groupColumn = "Group",
    groupLevels = c("COVID Moderate", "COVID Severe"),
    candidateEdgeTable = measuredPriorNetwork,
    correlationMethod = "pearson",
    minimumAbsoluteCorrelation = 0.3,
    adjustedPValueThreshold = 0.05,
    pAdjustMethod = "fdr",
    edgeWeightMethod = "signedZScore",
    scoreColumn = "rawRewiringScore",
    nPermutations = 99,
    seed = 1,
    keepPermutationScores = TRUE,
    resultName = "differentialSparseRewiringPermutation",
    storeResult = TRUE)


rewiringPermutation <-
    validationResults(analysisCovid)[["differentialSparseRewiringPermutation"]]

rewiringPermutation$inferenceStatus

rankedRewiring <- rewiringPermutation$rewiringTable
rankedRewiring <- rankedRewiring[
    order(
        rankedRewiring$adjustedPermutationTailProbability,
        -rankedRewiring$rawRewiringScore,
        na.last = TRUE
    ),
]

head(rankedRewiring, 20)

```

# Result export

Once results are generated, the network and node rewiring scores can be
supplied to `prepareCytoscapeTables()` to create tables that can be
imported into Cytoscape for further network analysis.

`writeNetworkTables()` writes the node and edge tables to a specified
directory. The vignette writes to `tempdir()` so package checks do not
write into the user's working directory.

```{r}
cytoscapeTables <- prepareCytoscapeTables(networkEdgeTable = differentialSparseNetwork,
                                          rewiringTable = rankedRewiring)

writeNetworkTables(cytoscapeTables,
                   directoryPath = file.path(tempdir(),
                                             "cornettoCovidSeverityNetwork"),
                   prefix = "covidModerateVsSevere",
                   fileFormat = "csv")
```

# Session information

```{r session-info}
sessionInfo()
```
