---
title: "CorNetto COVID-19 Severity Workflow"
author: "Bradley Ward"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{CorNetto COVID Severity Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(CorNetto)
extdataDir <- system.file("extdata", package = "CorNetto")
```

# Introduction

CorNetto builds knowledge-guided multi-omic correlation networks from
normalized assay data stored in a `MultiAssayExperiment`. It is intended
for workflows where transcriptomic, proteomic, metabolomic, or related
assays are already normalized and the analysis question is whether groups
differ in correlation structure, prior-supported connectivity, or
node-level rewiring.

The package uses standard Bioconductor containers for assay and sample
metadata, and returns standardized edge tables that can be combined,
converted to `igraph` objects, or exported as Cytoscape-ready node and
edge tables.

This vignette uses a small packaged subset of proteomic, transcriptomic,
and metabolomic data from moderate and severe COVID-19 patients. The subset
is intended to demonstrate the workflow during package checks; it is not
sized for biological discovery. The workflow first creates a prior knowledge
interactome, maps prior-guided differential correlations onto this structure,
and calculates node rewiring scores.

Because the sparse workflow uses a prior knowledge network to constrain
correlation calculations, it can demonstrate multi-omic network analysis
without running dense all-pairs correlations.

# Version information

```{r version-information}
c(
    R = R.version.string,
    Bioconductor = as.character(BiocManager::version()),
    CorNetto = as.character(utils::packageVersion("CorNetto"))
)
```

# Installation

After acceptance to Bioconductor, install CorNetto with:

```{r install, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}
BiocManager::install("CorNetto")
```

# Import prior knowledge networks into the environment

CorNetto is able to accept any kind of prior knowledge networks provided the 
columns can be mapped correctly from the data provided.

`readKnowledgeNetwork()` function takes a delimited table, optionally remaps
the input column names to a common convention, and standardizes it for downstream
CorNetto functions. The minimum columns required are `fromFeatureIdentifier`, 
`toFeatureIdentifier`, `fromAssayName `, and `toAssayName`. Other columns are 
optional but can be useful. The output is a standardized S4Vectors DFrame.

`combineKnowledgeNetworks()` takes these standardized networks and combines their rows while removing duplicate edges (optional).

```{r load prior data}

#Set column mapping
priorColumnMap <- c(
    fromFeatureIdentifier = "from",
    toFeatureIdentifier = "to",
    fromFeatureName = "fromName",
    toFeatureName = "toName",
    fromAssayName = "fromOmic",
    toAssayName = "toOmic",
    edgeType = "edgeType",
    edgeDirection = "edgeDirection",
    evidenceScore = "edgeWeight"
)
#Import prior data
stringNetwork <- readKnowledgeNetwork(filePath = file.path(extdataDir,
                                                           "priorNetworks",
                                                           "string.csv"),
                                      columnMapping = priorColumnMap,
                                      knowledgeSource = "STRING")

tfNetwork <- readKnowledgeNetwork(filePath = file.path(extdataDir,
                                                       "priorNetworks",
                                                       "ttrust.csv"),
                                      columnMapping = priorColumnMap,
                                      knowledgeSource = "TRRUST")

mirnaNetwork <- readKnowledgeNetwork(filePath = file.path(extdataDir,
                                                          "priorNetworks",
                                                          "tarbase.csv"),
                                      columnMapping = priorColumnMap,
                                      knowledgeSource = "TarBase")

rnaProteinNetwork  <- readKnowledgeNetwork(filePath = file.path(extdataDir,
                                                                "priorNetworks",
                                                                "translation.csv"),
                                      columnMapping = priorColumnMap,
                                      knowledgeSource = "RNA_to_protein")

dnaRnaNetwork  <- readKnowledgeNetwork(filePath = file.path(extdataDir,
                                                            "priorNetworks",
                                                            "transcription.csv"),
                                      columnMapping = priorColumnMap,
                                      knowledgeSource = "DNA_to_RNA")

metaboliteProteinNetwork <- readKnowledgeNetwork(filePath = file.path(extdataDir,
                                                                      "priorNetworks",
                                                                      "kegg_recon3D.csv"),
                                                 columnMapping = priorColumnMap,
                                                 knowledgeSource = "KEGG_Recon3D")
#Combine networks
combinedPriorNetwork <- combineKnowledgeNetworks(stringNetwork,
                                                 tfNetwork,
                                                 mirnaNetwork,
                                                 rnaProteinNetwork,
                                                 dnaRnaNetwork,
                                                 metaboliteProteinNetwork,
                                                 removeDuplicates = TRUE)
```

# Import experimental abundance data into the environment

Currently CorNetto is able to accept any normalised and fully preprocessed 
abundance or count matrix relating to proteomics, transcriptomics, or metabolomics.

`createAnalysisData()` creates a MultiAssayExperiment from these normalized assay objects and sample metadata. 

`filterSamples()` filters the MultiAssayExperiment based on a group column and a range of values.

```{r load experimental data}
#Import patient data
patientInformation <- read.csv(file.path(extdataDir,
                                         "covidData",
                                         "patientInformation.csv"))
rownames(patientInformation) <- patientInformation$Visit_ID

#Import omics data
dataList <- list(RNA = read.csv(file.path(extdataDir,
                                          "covidData",
                                          "rnaMatrix.csv"),
                                row.names = 1),
                 Protein  = read.csv(file.path(extdataDir,
                                               "covidData",
                                               "proteinMatrix.csv"),
                                     row.names = 1),
                 Metabolite  = read.csv(file.path(extdataDir,
                                                  "covidData",
                                                  "metaboliteMatrix.csv"),
                                        row.names = 1))

#Create analysis data
analysisData <- createAnalysisData(assayList = dataList,
                                   sampleData = patientInformation)

#Filter data for only COVID-19 patients at visit 1
analysisCovid <- filterSamples(analysisData,
                               groupColumn = "Group",
                               groupLevels = c("COVID Severe",
                                               "COVID Moderate"))
analysisCovid <- filterSamples(analysisCovid,
                               groupColumn = "Visit",
                               groupLevels = "Visit 1")
```

# QC checks of all imported objects

Following data importation, objects can be checked for structural conformity and
diagnostic summaries.

`validateAnalysisData()` checks the CorNetto analysis object for structural conformity, specifically:

* is this a MultiAssayExperiment?
* does it contain assays?
* are sample identifiers valid and unique?
* are assay columns matched to colData?
* are feature/sample names present and non-duplicated?
* are assays extractable as numeric matrices?
* are required feature metadata fields present?

`validateKnowledgeNetwork()` checks the CorNetto prior-knowledge edge table
for structural conformity, specifically:

* is this a data frame-like edge table?
* are the required edge columns present?
* are source and target feature identifiers present?
* are source and target assay names present?
* are missing optional columns filled with standard defaults?
* are edge type, edge direction, and knowledge source represented consistently?
* is the network converted to the standardized CorNetto edge-table format?

These functions also run quietly after most CorNetto functions to ensure
conformity of output.

```{r validateAnalysisData}
#Check analysis data
analysisCovid <- validateAnalysisData(analysisCovid)
#Check knowledge networks
combinedPriorNetwork <- validateKnowledgeNetwork(combinedPriorNetwork)
```

`summarizeAnalysisData()` asks "what does this analysis object look like,
and are there analysis risks?" rather than only asking whether the object
is structurally valid.
```{r summarizeAnalysisData}
summarizeAnalysisData(analysisCovid,
                      groupColumn = "Group")
```

The output reports the number of features per assay, the number of samples
per assay, and the number of samples in each selected group. In this small
packaged example there are 12 moderate COVID-19 samples and 12 severe
COVID-19 samples.

Zero-variance and near-zero-variance features should be removed before
correlation analysis. These can be removed with `removeZeroVariance = TRUE`
and `minimumVariance` in `filterFeatures()`.

```{r filterFeatures}
analysisCovid <- filterFeatures(
    analysisCovid,
    removeZeroVariance = TRUE,
    minimumVariance = 0.01
)

summarizeAnalysisData(analysisCovid)
```

For the sparse workflow, candidate edges are restricted to prior-network
edges whose endpoint features remain in the filtered analysis object.
This keeps the example fast and makes the tested edge universe explicit.

```{r restrict-prior-network}
measuredNodeKeys <- unlist(
    lapply(
        names(MultiAssayExperiment::experiments(analysisCovid)),
        function(assayName) {
            assayObject <- MultiAssayExperiment::experiments(analysisCovid)[[assayName]]
            paste(
                assayName,
                rownames(SummarizedExperiment::assay(assayObject)),
                sep = "::"
            )
        }
    ),
    use.names = FALSE
)

measuredPriorNetwork <- filterNetworkByNodes(
    combinedPriorNetwork,
    nodes = measuredNodeKeys,
    mode = "both",
    nodeIdType = "key"
)

summarizeKnowledgeNetwork(measuredPriorNetwork, analysisCovid)
```

# Run prior-guided differential correlations

The main function for testing differential correlations between groups is 
`testDifferentialCorrelation`. Here we are testing differential correlations
for all the multi-omic candidate edges supported within our prior knowledge 
networks (edges to test supplied to the `candidateEdgeTable` argument).

The differential correlation result is added to `analysisCovid`, then
extracted with `differentialCorrelationResults()`. Alternatively,
`testDifferentialCorrelation()` can directly return the table with
`storeResult = FALSE`.
```{r testDifferentialCorrelation}
analysisCovid <- testDifferentialCorrelation(
    analysisData = analysisCovid,
    groupColumn = "Group",
    groupLevels = c("COVID Moderate", "COVID Severe"),
    candidateEdgeTable = measuredPriorNetwork,
    correlationMethod = "pearson",
    minimumAbsoluteCorrelation = 0.3,
    adjustedPValueThreshold = 0.05,
    pAdjustMethod = "fdr",
    resultName = "differentialSparseCorrelations"
)

differentialSparseCorrelations <-
    differentialCorrelationResults(analysisCovid)[["differentialSparseCorrelations"]]

as.data.frame(differentialSparseCorrelations)

```

# Build differential network.

The next step builds a differential correlation network from the
`differentialSparseCorrelations` table. The function returns the network
directly unless `storeResult = TRUE`, in which case `analysisData` and
`resultName` are supplied.

Here the resulting network is not filtered further. In a real analysis,
`minimumAbsoluteCorrelation` and `differenceAdjustedPValueThreshold` can
be tightened to remove weakly correlated or non-significant edges.
```{r}
analysisCovid <- createDifferentialCorrelationNetwork(
    differentialCorrelationTable = differentialSparseCorrelations,
    differenceAdjustedPValueThreshold = 1,
    minimumAbsoluteCorrelation = 0,
    edgeWeightMethod = "absoluteZScore",
    analysisData = analysisCovid,
    resultName = "differentialSparseNetwork",
    storeResult = TRUE)

differentialSparseNetwork <-
    differentialCorrelationNetworks(analysisCovid)[["differentialSparseNetwork"]]

as.data.frame(differentialSparseNetwork)

```

# Calculate rewiring scores.

Rewiring scores summarize the amount of differential-correlation signal
incident on each node in the differential network. Here the rewiring table
is stored in `analysisCovid` and extracted with `rewiringResults()`.

```{r calculateRewiringScores}
analysisCovid <- calculateRewiringScores(
    differentialCorrelationNetwork = differentialSparseNetwork,
    analysisData = analysisCovid,
    resultName = "differentialSparseRewiring",
    storeResult = TRUE
)

rewiringScores <- rewiringResults(analysisCovid)[["differentialSparseRewiring"]]

rewiringScores <- rewiringScores[
    order(rewiringScores$degreeMatchedZScore, decreasing = TRUE),
]

head(rewiringScores, 20)
```

# Run permutation validation.

`permuteRewiringScores()` permutes the supplied group labels, reruns the
differential correlation testing, rebuilds the networks, and calculates
empirical node-level rewiring p-values. This vignette uses only two
permutations so that package checks run quickly; real analyses should use
many more permutations.

```{r}
analysisCovid <- permuteRewiringScores(
    analysisData = analysisCovid,
    groupColumn = "Group",
    groupLevels = c("COVID Moderate", "COVID Severe"),
    candidateEdgeTable = measuredPriorNetwork,
    correlationMethod = "pearson",
    minimumAbsoluteCorrelation = 0.3,
    adjustedPValueThreshold = 0.05,
    pAdjustMethod = "fdr",
    differenceAdjustedPValueThreshold = 1,
    edgeWeightMethod = "absoluteZScore",
    scoreColumn = "rawRewiringScore",
    nPermutations = 2,
    seed = 1,
    keepPermutationScores = TRUE,
    resultName = "differentialSparseRewiringPermutation",
    storeResult = TRUE)


rewiringPermutation <-
    validationResults(analysisCovid)[["differentialSparseRewiringPermutation"]]

validatedRewiring <- rewiringPermutation$rewiringTable
validatedRewiring <- validatedRewiring[order(validatedRewiring$empiricalAdjustedPValue,
                                             -validatedRewiring$rawRewiringScore,
                                             na.last = TRUE),
                                       ]

head(validatedRewiring, 20)

```

# Result export

Once results are generated, the network and node rewiring scores can be
supplied to `prepareCytoscapeTables()` to create tables that can be
imported into Cytoscape for further network analysis.

`writeNetworkTables()` writes the node and edge tables to a specified
directory. The vignette writes to `tempdir()` so package checks do not
write into the user's working directory.

```{r}
cytoscapeTables <- prepareCytoscapeTables(networkEdgeTable = differentialSparseNetwork,
                                          rewiringTable = validatedRewiring)

writeNetworkTables(cytoscapeTables,
                   directoryPath = file.path(tempdir(),
                                             "cornettoCovidSeverityNetwork"),
                   prefix = "covidModerateVsSevere",
                   fileFormat = "csv")
```

# Session information

```{r session-info}
sessionInfo()
```

